Mirror, mirror on the wall. Which's the best match of them all?

### 2024-01-19 PNAS/index.Rnw
updated E language and family-wiseness

As the number of comparisons increases, the probability of encountering a coincidental match increases.
Statisticians call this the \textit{familywise error rate} ("family" = number of comparisons); it is an important quantity to control when conducting a series of tests.

@@

For a single-comparison FPR of $e$, the familywise FPR for $n$ comparisons is $1 - [1-e]^n$.
\Cref{fig:Accuracy} shows the familywise accuracy rate on the y-axis; if 90\% is an acceptable overall accuracy, 2 comparisons can be made when $e = 0.05$; this increases to 10 comparisons when $e =  0.01$.
Thus, at an error rate of \Sexpr{sprintf("e = %.2f", fpr)}, examiners can conduct approximately 5 comparisons without exceeding the familywise FPR of 10\%.

@@

Forensic practitioners often report the findings from their examinations in the form of a categorical conclusion reflecting a single decision.
This is misleading when the decision relies on multiple comparisons which are not individually presented in reports or testimony.
In this short contribution, we have shown that the implicit comparisons performed during forensic analysis of wire cuts increase the familywise error rate.
Omission of the underlying details is not necessarily due to malice; rather, it results from protocols which focus on finding the best possible match between two items and evaluating that match.

@@

First, examiners should report (or defense attorneys should request) the overall length or area of surfaces generated from blade cuts, test fires, and equivalent procedures intended to assess the marks made by a tool submitted for comparison with evidence.
Examiners should also report the total consecutive length or area of the recovered evidence. These pieces of information will allow for calculations of minimum and maximum number of comparisons (as demonstrated in this paper), facilitating calculation of overall error rates.

Second, researchers should conduct studies relating the length/area of comparison surface to the error rate. For instance, we have pooled studies looking at bullet striations and firing pin shear marks. The striated surfaces are of orders of magnitude different lengths, but they represent our best estimate of the error rate for striated comparisons.
This was necessary because we could not find black-box error rate studies of wire cuts.
These studies should be Type II studies as defined by Koehler \citep{koehlerIntuitiveErrorRate2017a}, in that they should be designed to assess the actual error rate when examiners are making difficult comparisons.

Finally, when databases are used at any stage of the forensic evidence evaluation process (from suitability assessment and triage to reports which will be used at trial), the number of database items searched (or comparisons made) and the number of results returned must be reported. Furthermore, the number of results used for further manual comparison should also be reported. For example, if a firearms examiner searches a local NIBIN database with 1000 entries, requests the 20 closest matches to her evidence, and then carries out a physical examination of five exemplars from the list of 20, all of those values should be clearly reported to enable calculation of the potential FPR.
This will help make the multiple comparison issue accessible to everyone involved in evaluating the value of forensic evidence, from examiners to lawyers to jurors and judges.

>>>>

As the number of comparisons increases, the probability of encountering a coincidental match increases.
Statisticians call this the \textit{family-wise error rate} $E$; it is an important quantity to control when conducting a series ("family") of tests.

@@

For a single-comparison FPR of $e$, the family-wise FPR for $n$ comparisons, $E_n$ is $1 - [1-e]^n$.
\Cref{fig:Accuracy} shows the family-wise accuracy rate on the y-axis; if 90\% is an acceptable overall accuracy, 2 comparisons can be made when $e = 0.05$; this increases to 10 comparisons when $e =  0.01$.
Thus, at an error rate of \Sexpr{sprintf("e = %.2f", fpr)}, examiners can conduct approximately 5 comparisons without exceeding the family-wise FPR of 10\%.

@@

Forensic practitioners often report the findings from their examinations in the form of a categorical conclusion reflecting a single decision.
This is misleading when the decision relies on multiple comparisons which are not individually presented in reports or testimony.
In this short contribution, we have shown that the implicit comparisons performed during forensic analysis of wire cuts increase the family-wise error rate.
Omission of the underlying details is not necessarily due to malice; rather, it results from protocols which focus on finding the best possible match between two items and evaluating that match.

@@

First, examiners should report (or defense attorneys should request) the overall length or area of surfaces generated during the examination process.
% from blade cuts, test fires, and equivalent procedures intended to assess the marks made by a tool submitted for comparison with evidence.
Examiners should also report the total consecutive length or area of the recovered evidence.
These pieces of information will take the place of $\ell$ and $d$ and facilitate calculation of examination-wide error rates.

Second, researchers should conduct studies relating the length/area of comparison surface to the error rate.
For instance, we have pooled studies looking at bullet striations and firing pin shear marks.
The striated surfaces are of orders of magnitude different lengths, but they represent our best estimate of the error rate for striated comparisons.
This was necessary because we could not find black-box error rate studies of wire cuts.
New studies should be Type II studies as defined by Koehler \citep{koehlerIntuitiveErrorRate2017a}, designed to assess the actual error rate when examiners are making difficult comparisons.

Finally, when databases are used at any stage of the forensic evidence evaluation process (from suitability assessment and triage to reports which will be used at trial), the number of database items searched (or comparisons made) and the number of results returned must be reported.
Additionally, the number of results used for further manual comparison should also be reported.
For example, if a firearms examiner searches a local NIBIN database with 1000 entries, requests the 20 closest matches to her evidence, and then carries out a physical examination of five exemplars from the list of 20, all of those values should be clearly reported to enable estimation of the examination-wise error rate.
This will help make the multiple comparison issue accessible to everyone involved in evaluating the value of forensic evidence, from examiners to lawyers to jurors and judges.
